Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences

### 2016-09-19 paper/paper.tex
shared y-axis for R and M only results, small overhaul for paper up to section 3

\subsection{Languages and linguistic behavior}
Lexica are taken to codify the truth-conditions of a language's expressions, i.e., its semantics. Given a state of affairs and a lexicon, language users can judge whether an expression is true or false. A convenient way to represent such lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs and $M$ the set of messages in the lexicon \citep{franke+jaeger:2014}. For instance, the following two lexica fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$:

@@

In words, according to lexicon $L_a$, $m_1$ is true of $s_1$ as well as $s_2$. In contrast, message $m_1$ is only true of $s_1$ in $L_b$. Otherwise, the two languages are truth-conditionally equivalent.

To make the distinction betweeen semantics and pragmatics precise, we distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce and interpret messages literally. That is, their linguistic choices are solely guided by their lexica. In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices. For instance, given that $m_2$ is unambiguously associated with state $s_1$, a rational speaker of $L_a$ who reasons about her addressee should use $m_1$ to signal state $s_2$. Analogously, should rational hearers expect their interlocutors to reason along these lines, they will interpret the messages accordingly. Note in particular that according to this sketch of the pragmatic strenghtening of $m_1$, $L_a$ is indistinguishable from $L_b$ in terms of expressiveness.

Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic counterparts \citep{benz+etal:2005a,franke:2009,franke+jaeger:2014}, this kind of signaling behavior is captured by a hierarchy over reasoning types. The hierarchy's bottom, level $0$, corresponds to literal language use. Language users of level $n + 1$ behave rationally according to (expected) level $n$ behavior of their interlocutors. The behavior of literal level $0$ and pragmatic level $n+1$ hearers of a language $L$ is is captured by their respective selection functions in (\ref{h:level0}) and (\ref{h:leveln}). Mutatis mutandis for the speaker functions in (\ref{s:level0}) and (\ref{s:leveln}).

@@

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon, weighted by her prior over states, $pr \in \Delta(S)$. The latter plays an important role when hearers face ambiguous messages about which the prior is informative.

The literal speaker's choice in (\ref{s:level0}) is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\alpha$ increases, choices made in production are more rational; higher values lead to more deterministic expected utility maximizing behavior.

Pragmatic behavior resembles its literal counterpart. As described above, the crucial difference is that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior. That is, they reason about how a rational level $n$ interlocutor would use or interpret a message and behave according to these expectations. Additionally, pragmatic production is further regulated by a parameter $\alpha$ which controls the tension between semantics and pragmatics, $\alpha \in (0,1]$. Lower values lead to more literal production, whereas higher values lead to more pragmatic behavior.

The combination of a lexicon with its use, i.e., a degree of pragmatic sophistication, yields a type $t \in T$. Accordingly, $T$ contains all all lexica and reasoning level pairings under consideration. Types are the basic units on which our population dynamics operate.

\subsection{Replication \& expressivity}
On a population level, expressiveness, or communicative efficiency, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under this view, successful communication confers a higher fitness to some types relative to less successful ones. As a consequence, types with a higher fitness replicate and spread through the population, whereas the proportion of communicatively less efficient types decreases. This association of a type's communicative success within a population with changes in the types present in it creates a feedback loop that pressures the population towards greater expressivity. The replicator equation gives us the means to make these dynamics precise.

The proportion of types in a given population is captured by a vector $x$, where $x_i$ is type $i$'s proportion in the population. The fitness of a type $i$, $f_i$, is given by its expected utility when interacting in this population. That is, its fitness is the sum of its weighted expected communicative success, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$. The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$, and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$. $U_S(x,y)$ and $U_R(x,y)$ are respectively $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$ for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.\footnote{Note that the definition of $U_R(\cdot,\cdot)$ implies equal sender and receiver payoff in an interaction. This need not be so in the general case but suffices for our application.} Lastly, the average fitness of the population is captured by $\Phi$, $\Phi = \sum_i x_i f_i$. This term serves as a normalizing constant for the (discrete) replicator equation; $\dot{x}_i = \frac{x_i f_i}{\Phi}$

Under its biological interpretation, the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations. In analogy to biological replication, many aspects of natural language are subject to processees of change across varied time-spans. For example, the replicator equation can be understood as a learning across generations as e.g. in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation (see \citealt[\S3.3]{benz+etal:2005b} for discussion). In the following, we take the latter view and assume that interlocutors adapt their language's semantics and how it is used to that which works best within their population. In contrast to the effect of language acquisition from a generation to the next, this means that expressivity exherts its pressure on the current members of a linguistic community.

Nowak \& et al. did not only consider replication, but also recognized the relevance of variation introduced by a language's transmition across generations, construed as mutation. As a consequence, the offspring of a type may end up adopting a different type than their parents. However, in this work mutation rates were construed as independent from a type. That is, the variation introduced by generational turnovers did not depend on a the relative learnability of a type based on learning data or learning biases. Therefore, to address the issue of selecting a particular type over (near) functional equivalents, we turn to a different strand of research in cultural evolution: {\em iterated learning}.

\subsection{Mutation \& learning}
Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on. In linguistic terms, this process can be construed by chains of parents and children, where the parent produces linguistic data from which the child infers a language. The latter then goes on to produce linguistic data for a new generation of na\"ive learners. Following \citet{griffiths+kalish:2007} we model learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases. They then select a type from the resulting posterior distribution according to their learning strategy.

A central result of iterated learning is that this it leads to simpler and more regular languages due to the pressure towards greater learnability it exherts (surveys of experimental data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}). Importantly, experimental and mathematical results suggest the results of this process to reflect the learners' learning biases, codified as a prior $P \in \Delta(\mathcal{T})$. This bias can be thought of as the amount of data a learner would require in order to adopt a language or, in our case, a combination of a lexicon and a signaling behavior (cf. \citealt[450]{griffiths+kalish:2007}).

However, the extent of the influence of the prior on the dynamics was shown to depend on assumptions about the strategy employed by learners. While simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}, the mathematical characterization provided by \citet{griffiths+kalish:2007} showed that, instead, iterated learning converged to the prior. That is, the distribution over languages in a population or, from an individual's perspective, the likelihood of learning a language corresponded to the learners' prior distribution -- irrespective of the amount of input given to learners. The divergence between these predictions crucially depends on the assumed learning strategy. On the one extreme, Griffith \& Kalish's prediction holds for learners that sample from the posterior, on the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence.

In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live in the range between posterior sampling and MAP, $l \geq 1$. When $l = 1$ learners sample from the posterior. As $l$ increases towards infinity, the learners tendency to posterior maximization increases.

More generally, we combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$, where $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$. This, in turn, is proportional to the probability of $i$ producing the learning data and that of $j$ given the data. That is, the learner's task is to perform a joint inference over lexica and signaling behavior.

@@

\subsection{Summary}
So far, we have argued for expressivity, learnability and simplicity as central pressures that apply on the cultural evolution of language. The dynamics we propose model them as fitness-relative replication, mutation based on iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings, respectively. Taken together the dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

@@

The basic units that these pressures apply to are a combination of semantics, i.e., a lexicon, and its use. The expressivity of these types depends on their communicative efficiency within a population. Their learnability depends on the fidelity by which they can be infered by a new generation of learners based on production data. In particular, learners perform a joint inference on types of linguistic behavior and lexical meaning. This process is influenced by the learners' preference for simpler lexical representations, codified as a learning bias.

@@

\section{Lack of semantic upper-bounds in lexical meaning}
We set out to investigate the prevalence of lexical meanings that allow for regular pragmatic enrichments over other potential alternatives. A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures} (cf. \citealt{horn:1972,gazdar:1979}). These inferences are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.

For instance, {\em some} is entailed by {\em all}; if it were true that `All students came to class', it would also be true that `Some students came to class'. However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not necessarily what their use is taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold. That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative ({\em all}), this alternative must not hold. In this way, `Some students came to class' is strengthened as conveying `Some but not all students came to class'. Analogously, a speaker can rely on her interlocutor to draw this inference without having to express this upper-bound overtly, e.g. by stating {\em some but not all}. Put differently, mutual reasoning about rational language use supplies a bound that rules out stronger alternatives pragmatically.

Note that this corresponds to our previous description of the pragmatic use of lexicon $L_a$, repeated below for convenience. A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ given that the latter is unambiguously associated with the latter state. The strength of this association being dependent on their degree of rationality $\lambda$ and their prior over states. Conversely, a pragmatic speaker will reason about her interlocutors interpretation and use the messages accordingly.

@@

We can now rephrase our initial question in terms of scalar implicatures: Why is the lack of lexical upper-bounds in weak scalar alternatives, as $m_1$ in $L_a$, regularly selected for over the alternative of lexicalizing it as in $L_b$? More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}?  This question is particularly striking considering the number of classes of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).

@@

%\begin{figure}
%\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/07fitness}
%\end{figure}
%
%\begin{figure}
%\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/07learning}
%\end{figure}

>>>>

\subsection{Languages and linguistic behavior}
Lexica codify the truth-conditions of a language's expressions, i.e., its semantics. A convenient way to represent such lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs, or meanings, to convey and $M$ a set of messages of the language \citep{franke+jaeger:2014}. For instance, the following two lexica fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$:

@@

In words, according to lexicon $L_a$, $m_1$ is true of state $s_1$ as well as of $s_2$. In contrast, message $m_1$ is only true of $s_1$ in $L_b$. Otherwise, the two languages are truth-conditionally equivalent.

To make the distinction betweeen semantics and pragmatics precise, we distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce and interpret messages literally. That is, their linguistic choices are guided by their lexica only. In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices. For instance, a rational speaker of $L_a$ who reasons about her addressee should use $m_1$ to signal state $s_1$ given that $m_2$ is already unambiguously associated with state $s_2$. Analogously, should rational hearers expect their interlocutors to reason along these lines, they will interpret ambiguous $m_1$ accordingly. Note in particular that according to this strenghtening of $m_1$, $L_a$ is indistinguishable from $L_b$ in terms of expressiveness if its users are pragmatic reasoners.

Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic counterparts \citep{benz+etal:2005a,franke:2009,franke+jaeger:2014}, this kind of signaling behavior is captured by a hierarchy over reasoning types. The hierarchy's bottom, level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$ behave rationally according to (expected) level $n$ behavior of their interlocutors. The behavior of literal and pragmatic hearers of a language $L$ is given by their respective selection functions in (\ref{h:level0}) and (\ref{h:leveln}). Mutatis mutandis for the speaker functions in (\ref{s:level0}) and (\ref{s:leveln}).

@@

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. The literal speaker's choice in (\ref{s:level0}) is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices made in production are more rational in that higher values lead to more deterministic in line with expected utility maximization.

For the most part, pragmatic behavior mirrors its literal counterpart. As described above, their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon. That is, they reason about how a rational level $n$ interlocutor would use or interpret a message and behave according to these expectations. Additionally, pragmatic production is further regulated by a parameter $\alpha$ which controls the tension between semantics and pragmatics, $\alpha \in (0,1]$. Lower values lead to more literal production, whereas higher values lead to stronger pragmatic behavior.

The combination of a lexicon with its use, i.e., a particular level of linguistic sophistication, yields a type $t$. Types are the basic units on which our population dynamics operate.

\subsection{Replication \& expressivity}
Communicative efficiency, or expressiveness, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under this view, a type's success in communication confers it a higher fitness relative to less successful ones. As a consequence they replicate more than other types, increasing their proportion in the population. This association of a type's communicative success within a population with changes in the types present in it creates a feedback loop that pressures the population towards greater expressivity. The replicator equation gives us the means to make these dynamics precise.

The proportion of types in a given population is captured by a vector $x$, where $x_i$ is type $i$'s proportion in the population. The fitness of a type $i$, $f_i$, is given by its expected utility in this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$. That is, its fitness is the sum of its expected communicative success with other types weighted by the latter type's population share. The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$, and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$. $U_S(x,y)$ and $U_R(x,y)$ are respectively $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$ for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.\footnote{Note that the definition of $U_R(\cdot,\cdot)$ implies equal sender and receiver payoff in an interaction. This need not be so in the general case but suffices for our application.} This quantity is symmetric, reflecting the probability of two types' mutual understanding. Lastly, the average fitness of the population is captured by $\Phi$, $\Phi = \sum_i x_i f_i$. This term serves as a normalizing constant for the (discrete) replicator equation; $\dot{x}_i = \frac{x_i f_i}{\Phi}$

Under its biological interpretation, the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations. In analogy to this kind of replication, many aspects of natural language are subject to processees of transmission and change across varied time-spans. For example, the replicator equation can be understood as a learning across generations as e.g. in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation (see \citealt[\S3.3]{benz+etal:2005b} for discussion). In the following, we take the latter view in assuming that interlocutors adapt their lexica and their use to that which works best within their population. It should be stressed, however, that the model itself is compatible with either view.

Nowak et al. did not only consider replication, but also recognized the important role of the variation that is introduced by a language's transmition across generations, construed as mutation. Due to this process, the offspring of a type may end up adopting a different type than that of its parent. Crucially, in this work mutation rates were modelled as begin independent from a type. This means that the variation introduced by generational turnovers did not depend on factors such as the relative learnability of a type. To address the issue of selecting a particular type over (near) functional equivalents, we turn to a different strand of research in cultural evolution: {\em iterated learning}.

\subsection{Mutation \& learning}
Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on. For linguistic purposes this process can be thought of as chains of parents and children, where the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners. Following \citet{griffiths+kalish:2007} we model learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases. They then select a type to adopt from the resulting posterior distribution.

Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (surveys of empirical data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}). Importantly, experimental and mathematical results suggest the results of this process to reflect learners' learning biases, codified in the following as a prior $P \in \Delta(\mathcal{T})$. A way to think about this bias is as the amount of data a learner would require in order to adopt a language -- or, in our case, a combination of a lexicon and a signaling behavior (cf. \citealt[450]{griffiths+kalish:2007}). Crucially, the extent of the prior's influence has been shown to strongly depend on the learning strategy assumed to underly the inference process. While simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}, the mathematical characterization provided by \citet{griffiths+kalish:2007} showed that, instead, iterated learning converged to the prior. That is, the distribution over languages in a population or, from an individual's perspective, the likelihood of learning a language corresponds to the learners' prior distribution, irrespective of the amount of input given to learners. This divergence in predictions can be traced back to differences in the selection of hypotheses from the posterior. On the one extreme, Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior. On the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live in the range between posterior sampling and MAP, $l \geq 1$. When $l = 1$ learners sample from the posterior. As $l$ increases towards infinity, the learners' tendency maximize the posterior increases.

More generally, we combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$. Just as in standard mutator dynamics, $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$. However, to make this process depend on a type's learnbility, this quantity is proportional to the probability of $i$ producing the learning data and that of $j$ given the data.

@@

\subsection{Summary}
So far, we have argued for expressivity, learnability and simplicity as central pressures that apply on the cultural evolution of language. These are modelled as fitness-relative replication, mutation based on iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings, respectively. Taken together the dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

@@

The basic units that these pressures apply to are a combination of semantics, i.e., a lexicon, and its use. Their expressivity depends on their communicative efficiency within a population, whereas their learnability depends on the fidelity by which they can be infered by a new generation of learners. Importantly, the latter is influenced by the learners' preference for simpler lexical representations, codified as a learning bias.

@@

\section{Lack of semantic upper-bounds in lexical meaning}
With this model, we set out to investigate the prevalence of lexical meanings that allow for regular pragmatic enrichments over other alternatives. A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures}. These inferences are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order. For instance, {\em some} is entailed by {\em all}; if it were true that `All students came to class', it would also be true that `Some students came to class'. However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not necessarily what their use is taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative, e.g. {\em all}, this alternative must not hold. In this way, `Some students came to class' is strengthened to convey `Some but not all students came to class'. Analogously, a speaker can rely on her interlocutor to draw this inference without having to express this upper-bound overtly, e.g. by stating {\em some but not all}. In other words, mutual reasoning about rational language use supplies a bound that rules out stronger alternatives pragmatically.

This corresponds to our previous description of the pragmatic use of lexicon $L_a$, repeated below for convenience. A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ given that the latter is unambiguously associated with $s_2$. The strength of this association dependens on the individuals' degree of rationality $\lambda$ and their prior over states. Conversely, a pragmatic speaker will reason about her interlocutor's interpretation and use the messages accordingly.

@@

Our initial question can now be rephrased in terms of scalar implicatures by asking for justifiations for the lack of lexical upper-bounds in weak scalar alternatives. That is, why semantics such as those of message $m_1$ in $L_a$ are regularly selected for over the alternative of lexicalizing it as in $L_b$. More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}?  This question is particularly striking considering the number of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).

@@

\begin{figure}
\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{./only-R-or-M}
\end{figure}
%
%\begin{figure}
%\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/07learning}
%\end{figure}
